---
title: "Climate Projections"
author: "Natalie Smith & Madi Calbert"
format: 
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
theme: Litera
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#setup

#load libraries:
library(lubridate)
library(ggplot2)
library(tidyverse)
library(chron)
library(ncdf4)
library(RColorBrewer)
library(lattice)
library(abind)
library(here)
```

1.  Select a sample of **2-3 different climate models and 2 different future scenarios** to work with.

    -   canESM5 ssp126 and ssp585

    -   MIROC6 ssp126 and ssp585

    -   Briefly summarize the choices of model and scenario you made and the reasons (note: you don’t have to go into a lot of detail here). (Natalie)

2.  Load the monthly surface temperature data from these models and scenarios into the

    programming tool of your choice. Note: the name of the surface temperature variable may differ

    depending on where you’re downloading from; I recommend getting it from the CMIP6

    website, where it will be called “tas”.

```{r}
#path and filename for data

#CanESM5 SSP126:
CAN_126 <- here("data", "tas_Amon_CanESM5_ssp126_r19i1p2f1_gn_201501-210012.nc")
tas <- "tas"  # this is the name of the variable you want to look at

CAN_126 <- nc_open(CAN_126)
print(CAN_126)

# ------------------- CAN_ESM5 585 ------------------- 
CAN_585 <- here("data", "tas_Amon_CanESM5_ssp585_r15i1p1f1_gn_201501-210012.nc")
tas <- "tas"  # this is the name of the variable you want to look at

CAN_585 <- nc_open(CAN_585)
print(CAN_585)

#--------

#MIROC6 SSP126:
MIROC_126 <- here("data", "tas_Amon_MIROC6_ssp126_r3i1p1f1_gn_201501-210012.nc")
tas <- "tas"  # this is the name of the variable you want to look at

MIROC_126 <- nc_open(MIROC_126)
print(MIROC_126)

#MIROC6 SSP585:
MIROC_585 <- here("data", "tas_Amon_MIROC6_ssp585_r1i1p1f1_gn_201501-210012.nc")
tas <- "tas"  # this is the name of the variable you want to look at

MIROC_585  <- nc_open(MIROC_585 )
print(MIROC_585 )

```

```{r}
# "actually read in the data" using ncvar_get; from the data get latitude and longitude

# ------------------- CAN_126 ------------------- 
lon <- ncvar_get(CAN_126 ,"lon") #longitude
nlon <- dim(lon) #sets dimensions of object
lat <- ncvar_get(CAN_126,"lat") #latitude
nlat <- dim(lat) #sets dimensions of object

head(lat)
head(lon)

# ------------------- CAN_585 ------------------- 
lon_c585 <- ncvar_get(CAN_585,"lon") #longitude
nlon_c585 <- dim(lon_c585) #sets dimensions of object
lat_c585 <- ncvar_get(CAN_585,"lat") #latitude
nlat_c585 <- dim(lat_c585) #sets dimensions of object

head(lat_c585)
head(lon_c585)

# ------------------- MIROC_126 -------------------
lon_M126 <- ncvar_get(MIROC_126,"lon") #longitude
nlon_M126 <- dim(lon_M126) #sets dimensions of object
lat_M126 <- ncvar_get(MIROC_126,"lat") #latitude
nlat_M126 <- dim(lat_M126) #sets dimensions of object

head(lat_M126)
head(lon_M126)

# ------------------- MIROC_585 -------------------
lon_M585 <- ncvar_get(MIROC_585,"lon") #longitude
nlon_M585 <- dim(lon_M585) #sets dimensions of object
lat_M585 <- ncvar_get(MIROC_585,"lat") #latitude
nlat_M585 <- dim(lat_M585) #sets dimensions of object

head(lat_M585)
head(lon_M585)

```

```{r}
# from the data get time

# ------------------- CAN_126 ------------------- 
time <- ncvar_get(CAN_126,"time")
tunits <- ncatt_get(CAN_126,"time","units") # what units does the time varibale use?
nt <- dim(time)

print(tunits)
#$hasatt [1] TRUE
#$value [1] "days since 1850-01-01 0:0:0.0"
#They are counted since the date January 1, 1850 at midnight (0:0:0.0)


# ------------------- CAN_585 ------------------- 
time_c585 <- ncvar_get(CAN_585,"time")
tunits_c585 <- ncatt_get(CAN_585,"time","units") # what units does the time varibale use?
nt_c585 <- dim(time_c585)

print(tunits_c585)

# ------------------- MIROC_126 -------------------
time_M126 <- ncvar_get(MIROC_126,"time")
tunits_M126 <- ncatt_get(MIROC_126,"time","units") # what units does the time variable use?
nt_M126 <- dim(time_M126)

print(tunits_M126)

# ------------------- MIROC_585 -------------------
time_M585 <- ncvar_get(MIROC_585,"time")
tunits_M585 <- ncatt_get(MIROC_585,"time","units") # what units does the time variable use?
nt_M585 <- dim(time_M585)

print(tunits_M585)

```

```{r}
# split into parts 

# ------------------- CAN_126 ------------------- 
tustr <- strsplit(tunits$value, " ") 
tdstr <- strsplit(unlist(tustr)[3], "-")
tmonth <- as.integer(unlist(tdstr)[2])
tday <- as.integer(unlist(tdstr)[3])
tyear <- as.integer(unlist(tdstr)[1])
rtime_canesm5 <- chron(time,origin=c(tmonth, tday, tyear)) #chronological order

# ------------------- CAN_585 ------------------- 
tustr_c585 <- strsplit(tunits_c585$value, " ") 
tdstr_c585 <- strsplit(unlist(tustr_c585)[3], "-")
tmonth_c585 <- as.integer(unlist(tdstr_c585)[2])
tday_c585 <- as.integer(unlist(tdstr_c585)[3])
tyear_c585 <- as.integer(unlist(tdstr_c585)[1])
rtime_c585 <- chron(time_c585,origin=c(tmonth_c585, tday_c585, tyear_c585)) #chronological order

# ------------------- MIROC_126 -------------------
tustr_M126 <- strsplit(tunits_M126$value, " ") 
tdstr_M126 <- strsplit(unlist(tustr_M126)[3], "-")
tmonth_M126 <- as.integer(unlist(tdstr_M126)[2])
tday_M126 <- as.integer(unlist(tdstr_M126)[3])
tyear_M126 <- as.integer(unlist(tdstr_M126)[1])
rtime_M126 <- chron(time_M126,origin=c(tmonth_M126, tday_M126, tyear_M126)) #chronological order

# ------------------- MIROC_585 -------------------
tustr_M585 <- strsplit(tunits_M585$value, " ") 
tdstr_M585 <- strsplit(unlist(tustr_M585)[3], "-")
tmonth_M585 <- as.integer(unlist(tdstr_M585)[2])
tday_M585 <- as.integer(unlist(tdstr_M585)[3])
tyear_M585 <- as.integer(unlist(tdstr_M585)[1])
rtime_M585 <- chron(time_M585,origin=c(tmonth_M585, tday_M585, tyear_M585)) #chronological order

```

```{r}
# ------------------- CAN_126 ------------------- 
# read in the temperature data: 
TS <- ncvar_get(CAN_126, "tas") # get the surface temp variable
dlname <- ncatt_get(CAN_126,tas,"long_name") # get the long name of the variable (tas)
dunits<- ncatt_get(CAN_126,tas,"units") # get the units of tas
fillvalue <- ncatt_get(CAN_126,tas,"_FillValue") # get the fill value --> value used to fill in values where there is no data

# ------------------- CAN_585 ------------------- 
TS_c585 <- ncvar_get(CAN_585, "tas") # get the surface temp variable
dlname_c585 <- ncatt_get(CAN_585,tas,"long_name") # get the long name of the variable (tas)
dunits_c585 <- ncatt_get(CAN_585,tas,"units") # get the units of tas
fillvalue_c585 <- ncatt_get(CAN_585,tas,"_FillValue") # get the fill value --> value used to fill in values where there is no data

# ------------------- MIROC_126 -------------------
TS_M126 <- ncvar_get(MIROC_126, "tas") # get the surface temp variable
dlname_M126 <- ncatt_get(MIROC_126,tas,"long_name") # get the long name of the variable (tas)
dunits_M126 <- ncatt_get(MIROC_126,tas,"units") # get the units of tas
fillvalue_M126 <- ncatt_get(MIROC_126,tas,"_FillValue") # get the fill value --> value used to fill in values where there is no data

# ------------------- MIROC_585 -------------------
TS_M585 <- ncvar_get(MIROC_585, "tas") # get the surface temp variable
dlname_M585 <- ncatt_get(MIROC_585,tas,"long_name") # get the long name of the variable (tas)
dunits_M585 <- ncatt_get(MIROC_585,tas,"units") # get the units of tas
fillvalue_M585 <- ncatt_get(MIROC_585,tas,"_FillValue") # get the fill value --> value used to fill in values where there is no data

```

MADI does map 

```{r}
# process temperature data 
# lat x lon x time

# ------------------- CAN_126 ------------------- 
month <- 1 # month (January)
tmp_slice <- TS[,,month]-273.15 # convert Kelvin to Celsius

# levelplot (heatmap) of the slice
grid <- expand.grid(lon=lon, lat=lat) #values for lon (longitude) and lat (latitude) and creates a data frame (or tibble)
cutpts <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50) #vector defines cut points for the color scale in the plot
levelplot(tmp_slice ~ lon * lat, data=grid, at=cutpts, cuts=11, pretty=T, 
  col.regions=(rev(brewer.pal(10,"RdBu"))))


# ------------------- CAN_585 ------------------- 
month <- 1 # month (January)
tmp_slice_c585 <- TS_c585[,,month]-273.15 # convert Kelvin to Celsius

# levelplot (heatmap) of the slice
grid_c585 <- expand.grid(lon=lon_c585, lat=lat_c585) #values for lon (longitude) and lat (latitude) and creates a data frame (or tibble)
cutpts_c585 <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50) #vector defines cut points for the color scale in the plot
levelplot(tmp_slice_c585 ~ lon * lat, data=grid_c585, at=cutpts_c585, cuts=11, pretty=T, col.regions=(rev(brewer.pal(10,"RdBu"))))


# ------------------- MIROC_126 -------------------
month <- 1 # month (January)
tmp_slice <- TS_M126[,,month]-273.15 # convert Kelvin to Celsius

# levelplot (heatmap) of the slice
grid_M126 <- expand.grid(lon=lon_M126, lat=lat_M126) #values for lon (longitude) and lat (latitude) and creates a data frame (or tibble)
cutpts_M126 <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50) #vector defines cut points for the color scale in the plot
levelplot(tmp_slice ~ lon * lat, data=grid_M126, at=cutpts_M126, cuts=11, pretty=T, 
  col.regions=(rev(brewer.pal(10,"RdBu"))))

# ------------------- MIROC_585 -------------------
month <- 1 # month (January)
tmp_slice <- TS_M585[,,month]-273.15 # convert Kelvin to Celsius

# levelplot (heatmap) of the slice
grid_M585 <- expand.grid(lon=lon_M585, lat=lat_M585) #values for lon (longitude) and lat (latitude) and creates a data frame (or tibble)
cutpts_M585 <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50) #vector defines cut points for the color scale in the plot
levelplot(tmp_slice ~ lon * lat, data=grid_M585, at=cutpts_M585, cuts=11, pretty=T, 
  col.regions=(rev(brewer.pal(10,"RdBu"))))

```

3\. Consider a region of interest. This should be a relatively large region (think 500 miles/

1000km or so). Provide a brief description of your region and its present climate, and the

types of future impacts you expect to be important there. (Madi)

4\. Plot trends in surface temperature for your region, including all models and future scenarios

you identified in step 1.



Natalie does time series 

```{r}
# time series of regionally averaged data 

# ------------------- CAN_126 ------------------- 

# southern california lat/lon
lats=which(lat >= 32 & lat <= 35)
lons=which(lon >= 241 & lon <= 245)

# average over all values corresponding to each time over the region
tsavg_canesm5 <- apply(TS [lons,lats,],3,mean) # 3 = dimension over which to apply mean; 3 is for time

# plot the time series
clim <- data.frame(time=rtime_canesm5, tsavg_canesm5=tsavg_canesm5) 
yrclim = clim %>% group_by(year(rtime_canesm5)) %>% summarize(Tann=mean(tsavg_canesm5)) #group by year, and summarize to show annual average temperature
yrclim$dt = unique(year(rtime_canesm5)) #create a new column with the unique years

#plot the time series: 
c126_plot <-ggplot(yrclim, aes(dt, Tann-273.15))+geom_point()+labs(y="Southern CA Temperature", x="Year")+ geom_smooth(method="lm") + 
  theme_minimal()

# ------------------- CAN_585 ------------------- 
# southern california lat/lon
lats=which(lat >= 32 & lat <= 35)
lons=which(lon >= 241 & lon <= 245)

# average over all values corresponding to each time over the region
tsavg_canesm5 <- apply(TS_c585[lons,lats,],3,mean) # 3 = dimension over which to apply mean; 3 is for time

# plot the time series
clim <- data.frame(time=rtime_c585, tsavg_canesm5=tsavg_canesm5) 
yrclim_c585 = clim %>% group_by(year(rtime_c585 )) %>% summarize(Tann=mean(tsavg_canesm5)) #group by year, and summarize to show annual average temperature
yrclim_c585$dt = unique(year(rtime_c585)) #create a new column with the unique years

#plot the time series: 
c585_plot <-ggplot(yrclim_c585, aes(dt, Tann-273.15))+geom_point()+labs(y="Southern CA Temperature", x="Year")+ geom_smooth(method="lm") + 
  theme_minimal()


# ------------------- MIROC_126 -------------------
# southern california lat/lon
lats=which(lat_M126 >= 32 & lat_M126 <= 35)
lons=which(lon_M126 >= 241 & lon_M126 <= 245)

# average over all values corresponding to each time over the region
tsavg_M126 <- apply(TS_M126[lons,lats,],3,mean) # 3 = dimension over which to apply mean; 3 is for time

# plot the time series
clim <- data.frame(time=rtime_M126, tsavg_M126=tsavg_M126) 
yrclim_M126 = clim %>% group_by(year(rtime_M126)) %>% summarize(Tann=mean(tsavg_M126)) #group by year, and summarize to show annual average temperature
yrclim_M126$dt = unique(year(rtime_M126)) #create a new column with the unique years

#plot the time series: 
m126_plot <- ggplot(yrclim_M126, aes(dt, Tann-273.15))+geom_point()+labs(y="Southern CA Temperature", x="Year")+ geom_smooth(method="lm") + 
  theme_minimal()


# ------------------- MIROC_585 -------------------
# southern california lat/lon
lats=which(lat_M585 >= 32 & lat_M585 <= 35)
lons=which(lon_M585 >= 241 & lon_M585 <= 245)

# average over all values corresponding to each time over the region
tsavg_M585 <- apply(TS_M585[lons,lats,],3,mean) # 3 = dimension over which to apply mean; 3 is for time

# plot the time series
clim <- data.frame(time=rtime_M585, tsavg_M585=tsavg_M585) 
yrclim_M585 = clim %>% group_by(year(rtime_M585)) %>% summarize(Tann=mean(tsavg_M585)) #group by year, and summarize to show annual average temperature
yrclim_M585$dt = unique(year(rtime_M585)) #create a new column with the unique years

#plot the time series: 
m585_plot <-ggplot(yrclim_M585, aes(dt, Tann-273.15))+geom_point()+labs(y="Southern CA Temperature", x="Year")+ geom_smooth(method="lm") + 
  theme_minimal()

#patchwork plots together
library(patchwork)
combined_plot <- c126_plot + c585_plot + m126_plot + m585_plot + plot_layout(ncol = 2)
combined_plot
```

Then describe your results, addressing the following:

-   How large are the temperature trends over the 21st century (roughly, 2015-2100), and how statistically significant are they? (Natalie)

A larger slope indicates a larger temperature increase over time. The slope of the line is the change in temperature per year.

```{r}
#trend estimates
#CanESM5 SSP126
model <- lm(Tann ~ dt, data = yrclim)
summary(model) #slope = 2.455e+02

#CanESM5 SSP585
model_c585 <- lm(Tann ~ dt, data = yrclim_c585)
summary(model_c585) #slope = 1.110e+02

#MIROC6 SSP126
model_M126 <- lm(Tann ~ dt, data = yrclim_M126)
summary(model_M126) #slope = 2.790e+02

#MIROC6 SSP585
model_M585 <- lm(Tann ~ dt, data = yrclim_M585)
summary(model_M585) #slope = 1.879e+02
```

-   How do the trend estimates compare between different models? Between different scenarios? (Madi)

```{r}
#Between different models? 
#run a t-test comparing models across SSSP126
t_test_results_SSP126 <- t.test(
  yrclim$Tann, 
  yrclim_M126$Tann, paired = FALSE, var.equal = TRUE
)
t_test_results_SSP126 #p-value < 2.2e-16

#run a t-test comparing models across SSSP585
t_test_results_SSP585 <- t.test(
  yrclim_c585$Tann, 
  yrclim_M585$Tann, paired = FALSE, var.equal = TRUE
)
t_test_results_SSP585 #p-value = 2.486e-13

#Between different scenarios?
#run a t-test comparing scenarios across MIROC models
t_test_results <- t.test(
  yrclim_M126$Tann, 
  yrclim_M585$Tann, paired = FALSE, var.equal = TRUE
)
t_test_results #p-value = 1.174e-09

#run a t-test comparing scenarios across CanESM models
t_test_results_canesm5 <- t.test(
  yrclim_c585$Tann, 
  yrclim$Tann, paired = FALSE, var.equal = TRUE
)
t_test_results_canesm5 #p-value = 1.958e-07 
```

summarize the results in a two-page report including an introduction, figures/tables, and discussion.
